---
title: "Data wrangling - HELPrct"
subtitle: "Week 5 Activity"
author: "Shannon Curran and Sophie Massolas"
date: "15 February 2026"
output: 
  html_notebook:
    fig_height: 3
    fig_width: 5
---

```{r message=FALSE}
# Good practice - clear the working environment
rm(list = ls())

# Load required packages
library(tidyverse)  # for data verbs such as group-by() & summarise()
library(mosaicData) # includes the HELPrct data set 
library(esquisse)   # includes the esquisser() interactive plot builder seen in Week 4

# Load the HELPrct data from mosaicData package into our RStudio environment
data("HELPrct", package = "mosaicData")

# Use View(HELPrct) in the console to open a tab in RStudio and look at the raw data directly
```


## Data Description

*TASK 1: Write a couple of sentences describing the `HELPrct` data. What is the setting for the data? What are they about?* 

[hint: you can use the help in RStudio here since the data were loaded from an R package]




## Basic Summaries with `summarise()` 

*TASK 2: Produce an expression involving `summarize()` and `HELPrct` that will produce the following.*

1. *number of people (cases) in `HELPrct` study*
2. *combined total number of times in the past 6 months all the people in `HELPrct` entered a detox program (measured at baseline)* 
3. *mean time (in days) to first use of any substance post-detox for all the people in `HELPrct`*



```{r}

```




## Subgrouping with `group_by()` 

*TASK 3: Repeat Task 2 above, but calculate the results group-by-group.  Show your code and write a sentence or two about what you observe in the results for each of the following:* 

1. *males versus females*

```{r}
# males versus females

```


2. *homeless or not*

```{r}
# homeless or not

```

3. *substance*

```{r}
# substance

```

4. *break down the homeless versus housed further, by sex*

```{r}
# homeless versus housed broken down by sex

```

5. *break down the homeless versus housed further, by substance*

```{r}
# homeless versus housed broken down by substance

```



## TASK 4:

*Include one or more interesting plots of the data involving at least 3 variables per plot. Write a few sentences to explain the story that your plot tells about these data.  You can use one of the relationships that you studied in Task 2, or you can explore a different group of variables in the `HELPrct` that show something interesting.  You should use the interactive tool `esquisser` but you must copy/paste the `ggplot2` code to create the plot here (i.e., do not embed a screenshot or image file).*  


* Plots vary
* Almost any plot that you want to post here would be fine as long as it involves at least 3 variables. For example, 
    - a bivariate display like a scatterplot or line plot that uses color or facets for some third categorical variable.  
    - a single quantitative variable could be shown as a density or violin plot, and then color and faceting could incorporate two categorical variables.  
    - a combination of strategies above could easily incorporate 4 variables (2 quantitative; 2 categorical)
